Quantum-assisted unsupervised data clustering

### 2020-05-03 manuscript.tex
Fix qcircuit figure and caption
See #14

The SOFM is one of the most widely-used unsupervised learning methods used in various areas of modern science. 
It was first proposed by Kohonen as a self-organizing unsupervised learning algorithm which produces feature maps similar to those occurring in the brain \cite{solan2001}. 
The SOFM algorithm operates with a set of input objects, each represented by a $N$-dimensional vector, 
and describes a mapping from a higher-dimensional input space to a lower-dimensional map space.

The input dimensions are associated with the features, 
and the nodes in the grid (called cluster vectors) are assigned the $N$-dimensional vectors. 
The components of these vectors are usually called weights. 
Initially the weight components are chosen randomly. 
We then can train our SOFM adjusting the components through the learning process which occur in the two basic procedures of selecting a winning cluster vector and updating its weights (Fig.~\ref{fig:sofm_fitting}). 
More specifically, they consist of four step process: \begin{enumerate*}
\item selecting an input vector randomly from the set of all input vectors; 
\item finding a cluster vector which is closest to the input vector; 
\item adjusting the weights of the winning node in such a way that it becomes even closer to the input vector; 
\item repeating this process for many iterations until it converges.
\end{enumerate*}

After the winning cluster vector is selected, the weights of the vector are adjusted according to
%
\begin{equation}
	\vec w_{i+1} =
		\vec w_i
		+ \alpha\left(\vec{x} - \vec w_i\right).
    \label{eq:learning}
\end{equation}
%
The above expression can be interpreted according to: 
if a component of the input vector $\vec{x}$ is greater than the corresponding weight $ \vec{w}_i $, 
increase the weight by a small amount with the learning rate $\alpha$; 
if the input component is smaller than the weight, decrease the weight by a small amount. 
The larger the difference between the input component and the weight component, the larger the increment (decrement). 
Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way 
that ensures each move is following the current trends inferred from their distances to the input objects. 
A visualisation of this process is shown in Fig. \ref{fig:sofm_fitting}.

Usually the winning cluster vector is selected based on the Euclidean distance between an input vector and the cluster vectors. 
In our approach, we use the Hamming distance instead of the Euclidean distance to select the winning cluster vector. 
It allows us to use a simpler encoding of the classical information into the quantum state and use an effective procedure for the calculation of the Hamming distance on the quantum machine, 
such as to reduce the number of calculations in number of cluster vectors in comparison to the classical case.

@@

%%%%%
\begin{figure*}[t]
	\includegraphics[width=1.9\columnwidth]{qcircuit.png}
	\caption{
		Quantum circuit for the quantum parallelized Hamming distance calculation between all samples and clusters. 
		First,  Hadamard gates are applied in order to obtain a superposition of the sample and cluster registers. 
		Second, we encoded information about pairwise different qubits in a quantum state of the sample register with applying the CNOT gates. 
		Third, Hamming distance values are encoded in the amplitudes of superposition with the control phase rotation and Hadamard gates. 
		Finally, a quantum state of the sample register returned to the initial basis for information retrieval. 	
	} 
	\label{fig:qcircuit}
\end{figure*}
%%%%%

@@

We now introduce an optimized algorithm for calculating the matrix of Hamming distances \cite{trugenberger2001} between a sample vector and all cluster vectors, making use of quantum parallelism.

%This allows for a simple encoding of the classical information into a quantum register.

@@

\begin{align}
    | \psi_1 \rangle & = 
    \mathrm{CNOT} (Y,X)| \psi_0 \rangle \nonumber \\
		& =  
    \frac{1}{\sqrt{kl}} \sum_{i, j=1}^{k} 
    | d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \rangle 
    | y^{(1)}_j, \dots, y^{(n)}_j \rangle
    | 0 \rangle 
\end{align}
%
where $d^\alpha_{ij} = \mathrm{CNOT}(y^\alpha_i, x^\alpha_j)$, $\alpha = 1 \dots n$, and $i,j$ are the qubit indexes in the registers. 
At this stage of the computation the $\left| X \right\rangle$ no longer stores the input vectors,
instead it stores the information about pairwise different qubits between the input vector $\{X\}$ and cluster vector $\{Y\}$. 
Next, for each pair $\{X\}$ and $\{Y\}$, the accumulated information of all the differences is projected onto the amplitude of the superposed state. 
This is achieved by applying the Hadamard gate on auxiliary qubit, 
followed by a controlled phase gate on $\left| Xa \right\rangle$ defined as
%
\begin{equation}
    \label{eq:control_phase_rotation}
    R_{(X,a)}(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & e^{-i\phi}
    \end{pmatrix} .
    \quad \phi = \frac{\pi}{n}
\end{equation}
%
Finally, another Hadamard gate is applied on the ancilla qubit (see Fig. \ref{fig:qcircuit}).

After the first Hadamard on the ancilla qubit the state is
%
\begin{equation}
    \left| \psi_2 \right\rangle = H_a\left| \psi_1 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
    \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle
    \dfrac{(\left| 0 \right\rangle + \left| 1 \right\rangle)}{\sqrt{2}}  .
\end{equation}
%
Applying the controlled phase gate the state then becomes
%
\begin{multline}
    \left| \psi_3 \right\rangle = R_{(X,a)}\left(\dfrac{\pi}{n}\right)\left| \psi_2 \right\rangle
    \\ = \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k} 
				\left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle
        \\ + \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k}
        \exp\left(\dfrac{-i \pi}{n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
\\ \times        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle
\end{multline}
%
Applying another Hadamard on the ancilla qubit we obtain
%
\begin{multline}
    \left| \psi_4 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
		\\ \times
        \left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+ 
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This completes the step for projecting differences between pairs of $\{X\}$ and $\{Y\}$ onto the amplitude of the auxiliary qubit. 
The process is done in the $x$-basis, achieved by the surrounding Hadamard gates. 
There are two possible measurement outcomes of the auxiliary qubit.  
Each pair of $\{X\}$ and $\{Y\}$ forms a subspace of the Hilbert space, 
the controlled phase gate ensures to change amplitudes of those outcomes within this subspace depending on how different the spin configurations between $\{X\}$ and $\{Y\}$ are.

@@

We now show experimental results for a proof-of-concept demonstration of the algorithm introduced in the previous section, on a 16-qubit quantum computer provided by IBM Q Experience.  
We perform unsupervised data clustering for three sets of paper abstracts from three different fields: 
quantum physics \cite{qml0, qml1, qml2}, 
medicine \cite{med0, med1, med2} 
and biology \cite{bio0, bio1, bio2}. 
Each set consists of three random papers that focus on one of following topics: 
``Quantum Machine Learning'' (QML), 
``Cancer'' (MED) 
and ``Gene Expression'' (BIO). 
Abstracts were vectorized by the bag-of-words model in order to choose most defining words in each data set (see Fig.~\ref{fig:vectorized_sample}) \cite{mctear2016}.  
This model represents text as a multiset ``bag'' of its words taking into account only multiplicity of words. 
Preparing the bag-of-words we excluded the words that appear only in one abstract and more than in 4 abstracts and we also excluded the word ``level'' from consideration due to the frequent overlap between the clusters because it gives instabilities for both classical and quantum algorithms. 
We restricted our bag-of-word size to 9 of the most frequent words from the full bags-of-word  due to limitations of the number of qubits. 
After vectorizing and pre-processing  the data, the clusters are well separated with the Hamming distance.  
We observe that distances between the abstracts inside clusters are smaller than distances between the abstracts from different clusters, 
showing successful self-organization (Fig.~\ref{fig:vectorized_sample}).

>>>>

The SOFM is one of the most widely-used unsupervised learning methods used in various areas of modern science. 
It was first proposed by Kohonen as a self-organizing unsupervised learning algorithm which produces feature maps similar to those occurring in the brain \cite{solan2001}. 
The SOFM algorithm operates with a set of input objects, each represented by a $N$-dimensional vector, 
and describes a mapping from a higher-dimensional input space to a lower-dimensional map space, commonly a bi-dimensional map.

The input dimensions are associated with the features, 
and the nodes in the grid (called cluster vectors) are assigned the $N$-dimensional vectors. 
The components of these vectors are usually called weights. 
Initially the weight components are chosen randomly 
and topological distances between neurons given. 
We then can train our SOFM adjusting the components through the learning process which occur in the two basic procedures of 
selecting a winning cluster vector, also called the best matching unit (BMU), and updating its weights (Fig.~\ref{fig:sofm_fitting}). 
More specifically, they consist of four step process: 
\begin{enumerate*}
\item selecting an input vector randomly from the set of all input vectors; 
\item finding a cluster vector which is closest to the input vector (BMU); 
\item adjusting the weights of the BMU and neurons close to it on feature map in such a way 
	that these vectors  becomes even closer to the input vector; 
\item repeating this process for many iterations until it converges.
\end{enumerate*}

On a step $t$ when the BMU $w_{c}$ is selected, 
the weights of the BMU and its neigbours on feature map are adjusted according to
%
\begin{equation}
    \label{eq:learning}
	\vec{w_{i}}(t + 1) 
	= \vec{w_i}(t)
	+ \theta(c, i, t) \alpha(t) 
		\left(\vec{x}(t) - \vec{w_i}(t)\right),
\end{equation}
%
where  $\alpha(t)$ is the learning rate and $\theta(c, i, t)$ is the neighbor function,
which defines the topological neighbor neurons to be updated. 
Note that neighbor function depend on distance on feature map given initialy, 
but not distance metrics between vectors.

Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way 
that ensures each move is following the current trends inferred from their distances to the input objects. 
A visualisation of this process is shown in Fig. \ref{fig:sofm_fitting}.

In SOFM original version was designed to cluster real-valued data
and the winning cluster vector is selected based on the Euclidean distance between an input vector and the cluster vectors.
The paper deals with binary vectors clustering problem 
and for binary data the Hamming distance is more suitable \cite{appiah2009, santana2017}.
Using a known technique of encoding classical information into a quantum register, 
based on probabilistic quantum memories \cite{trugenberger2001},
we introduce an optimized algorithm for calculating the matrix of Hamming distances 
between each pair of binary vectors  of two sets taking advantage of quantum parallelism.

@@

%%%%%
\begin{figure}[t]
	\includegraphics[width=\columnwidth]{qcircuit.eps}
	\caption{
		Quantum circuit for the quantum parallelized Hamming distance calculating  between all pairs of binary vectors from two sets ${X}$ and ${Y}$ encoded \cite{trugenberger2001} in $X$ and $Y$ quantum registers   respectively.
		First, we encoded information about pairwise different qubits in a quantum state of the $X$-register with applying the CNOT gates. 
		Second, Hamming distance values are extracted into the amplitudes of superposition with the controled rotation around $z$-axis gate and Hadamard gates. 
		Finally, a quantum state of the $X$-register returned to the initial basis for information retrieval. 	
	} 
	\label{fig:qcircuit}
\end{figure}
%%%%%

@@

@@

\begin{align}
    | \psi_1 \rangle & = 
    \mathrm{CNOT} (Y,X)| \psi_0 \rangle \nonumber \\
		& =  
    \frac{1}{\sqrt{kl}} \sum_{i, j=1}^{k} 
    | d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \rangle 
    | y^{(1)}_j, \dots, y^{(n)}_j \rangle
    | 0 \rangle 
\end{align}
%
where $d^\alpha_{ij} = \mathrm{CNOT}(y^\alpha_i, x^\alpha_j)$, $\alpha = 1 \dots n$, and $i,j$ are the qubit indexes in the registers. 
At this stage of the computation the $\left| X \right\rangle$ no longer stores the input vectors,
instead it stores the information about pairwise different qubits between the input vector $\{X\}$ and cluster vector $\{Y\}$. 
Next, for each pair $\{X\}$ and $\{Y\}$, the accumulated information of all the differences is projected onto the amplitude of the superposed state. 
This is achieved by applying the Hadamard gate on auxiliary qubit, 
followed by a controlled rotation around $z$-axis gate on $\left| Xa \right\rangle$ defined as
%
\begin{equation}
    \label{eq:control_phase_rotation}
    R_{(X,a)}(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & e^{-i \frac \phi 2} & 0 & 0 \\
        0 & 0 & 1 & 0 \\
		0 & 0 & 0 & e^{i \frac \phi 2}
    \end{pmatrix} .
    \quad \phi = \frac{\pi}{n}
\end{equation}
%
Finally, another Hadamard gate is applied on the ancilla qubit (see Fig. \ref{fig:qcircuit}).

After the first Hadamard on the ancilla qubit the state is
%
\begin{equation}
    \left| \psi_2 \right\rangle = H_a\left| \psi_1 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
    \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle
    \dfrac{(\left| 0 \right\rangle + \left| 1 \right\rangle)}{\sqrt{2}}  .
\end{equation}
%
Applying the controlled rotation around $z$-axis gate the state then becomes
%
\begin{multline}
    \left| \psi_3 \right\rangle = R_{(X,a)}\left(\dfrac{\pi}{n}\right)\left| \psi_2 \right\rangle
    \\ = \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k} 
				\left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle
        \\ + \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k}
        \exp\left(\dfrac{-i \pi}{n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
\\ \times        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle
\end{multline}
%
Applying another Hadamard on the ancilla qubit we obtain
%
\begin{multline}
    \left| \psi_4 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
		\\ \times
        \left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+ 
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This completes the step for projecting differences between pairs of $\{X\}$ and $\{Y\}$ onto the amplitude of the auxiliary qubit. 
The process is done in the $x$-basis, achieved by the surrounding Hadamard gates. 
There are two possible measurement outcomes of the auxiliary qubit.  
Each pair of $\{X\}$ and $\{Y\}$ forms a subspace of the Hilbert space, 
the controlled rotation around $z$-axis gate ensures to change amplitudes of those outcomes within this subspace depending on how different the spin configurations between $\{X\}$ and $\{Y\}$ are.

@@

We now show experimental results for a proof-of-concept demonstration of the algorithm introduced in the previous section, on a 16-qubit quantum computer provided by IBM Q Experience.  
We perform unsupervised data clustering for three sets of paper abstracts from three different fields: 
quantum physics \cite{qml0, qml1, qml2}, 
medicine \cite{med0, med1, med2} 
and biology \cite{bio0, bio1, bio2}. 
Each set consists of three random papers that focus on one of following topics: 
``Quantum Machine Learning'' (QML), 
``Cancer'' (MED) 
and ``Gene Expression'' (BIO). 
Abstracts were vectorized by the bag-of-words\cite{weikang2016} model in order to choose most defining words in each data set (see Fig.~\ref{fig:vectorized_sample}) \cite{mctear2016}.  
This model represents text as a multiset ``bag'' of its words taking into account only multiplicity of words. 
Preparing the bag-of-words we excluded the words that appear only in one abstract and more than in 4 abstracts and we also excluded the word ``level'' from consideration due to the frequent overlap between the clusters because it gives instabilities for both classical and quantum algorithms. 
We restricted our bag-of-word size to 9 of the most frequent words from the full bags-of-word  due to limitations of the number of qubits. 
After vectorizing and pre-processing  the data, the clusters are well separated with the Hamming distance.  
We observe that distances between the abstracts inside clusters are smaller than distances between the abstracts from different clusters, 
showing successful self-organization (Fig.~\ref{fig:vectorized_sample}).
